Exploring Rural Shrink Smart Through Guided Discovery Dashboards

### 2022-07-19 SDSS Journal Paper/Paper-Draft.tex
First Draft Updates

---
title: Exploring Rural Shrink Smart Through Guided Discovery Dashboards
short-title:  
author:
  - first: Denise                 # first name
    first-init: D.              # initial of first name
    last: Bradford                  # last name
    marker: '1'               # identifier for the affiliation
    footnote-id: 1              # optional identifier for the footnote
    email: denise.bradford@huskers.unl.edu    # optional email address
  - first: Susan
    first-init: S.
    last: VanderPlas
    marker: '1'
    footnote-id: 1
footnote:
  - content: Corresponding author
    type: corresp               # set type for the corresponding author
    id: 1
affiliation:
  - institution: University of Nebraska - Lincoln
    prefix: Department of Statistics        # optional
    country: United States of America
    marker: 1
classoption:
  - letterpaper    # or a4paper, etc.
  - inpress        # or review
# - linenumber     # add line number for submission
preamble: >
  \usepackage{amsfonts,amsmath,amssymb,amsthm}
  \usepackage{booktabs}
  \usepackage{lipsum}
abstract: >
  Many small and rural places are shrinking. Interactive dashboards are the most common use cases for data visualization and context for exploratory data tools. In our paper, we will explore the specific scope of how dashboards are used in small and rural area to empower novice analysts to make data-driven decisions. Our framework will suggest a number of research directions to better support small and rural places from shrinking using an interactive dashboard design, implementation and use for the every day analyst.
keywords:
# alphabetical; excluding anything in the title already; lower case
  - Interactive Dashboards
  - Exploratory Data Analysis (EDA)
  - Parallel Coordinate Plots (PCP)
  - Guided Discovery Learning (GDL)
bibliography: jdsart-sample.bib
output: jds.rmd::pdf_article
---

# Abstract

@@

# Introduction

@@

Data collected from the Index of Relative Rurality (IRR) \url{https://www.ers.usda.gov/topics/rural-economy-population/rural-classifications/} were used in the SCC dashboard to help classify the towns. The Index of Relative Rurality (IRR) is a continuous, threshold-free, and unit-free measure of rurality. It is an alternative to the traditional discrete threshold-based classifications.The IRR ranges between 0 (low level of rurality, i.e., urban) and 1 (most rural). Four steps are involved in its design:
\begin{enumerate}
\item Identifying the dimensions of rurality: population size, density, remoteness, and built-up area.
\item Selecting measureable variables to adequately represent each dimension:
    \begin{itemize}
        \item Size: logarithm of population size
        \item Density: logarithm of population density.
        \item Remoteness: network distance.
        \item Built-up area: urban area (as defined by the US Census Bureau) as a percentage of total land area.
    \end{itemize}
\item Re-scaling the variables onto bounded scales that range from 0 to 1.
\item Selecting a link function: unweighted average of the four re-scaled variable.
\end{enumerate}

@@

One of the interesting features of this assembled dataset is that missing data can be missing for multiple reasons: not all state data is complete, but data about certain services may also be missing because towns do not offer that service.
Thus, in addition to the usual challenges of working with real-world data that is "messy" in a variety of ways, we also have to contend with missing data that is missing due to the size of the community or the lack of services. This makes both visualization and statistical analysis more complicated.

# Methods

@@

Our audience being small rural towns, traditionally, do not have access:
\begin{itemize}
     \item Data scientists 
      \item Ability to easily leverage data collected locally to support decisions 
\end{itemize}
Our research team will provide communities with data about services in small town Iowa in order to assist with developing strategies to improve quality of life for their residents amid shrinking populations.

While data is important for data-driven decisions, we need to help minimize the amount of cognitive load for the town-leaders to digest the amount of data will be conveyed, as a result we will create a central place for the excess of data to be conveyed in a single comprehensive visual. We will utilize Parallel Coordinate Plot (PCPs). Create a map feature that will allow for town leaders to understand the distance of essential QoL components (e.g. Fire Dept, Schools, etc.). Lastly, we will add an overview of town general basics related to the seven QoL factors.

@@

Thus, our initial set of dashboard design principles is as follows: 
\begin{itemize}
\item The town leaders are the focus audience; thus, the town itself should be the central focus of the app.
\item Facilitate comparisons with other towns in order to allow the user to explore other potential solutions to offering services that enhance resident quality of life.
\item Present the user with peer comparisons in order to widen the scope of exploration beyond the initial set of obvious peers in the local region.
\item Allow for more detailed data and feature requests to improve the dashboard design over time.
\end{itemize}

@@

\subsubsection{Dashboard Components}
The large set of publicly available data (primarily from \url{data.iowa.gov}) we have assembled is useful, but we must be careful with how we present this data because it would be easy to overwhelm the user with small details that mask the bigger picture. We select a small subset of towns (out of the 999 towns in Iowa) and a small subset of variables of interest to start with, and then allow the user to increase the complexity of the display in accordance with their interest. This avoids some of the pitfalls of dashboard design that can easily lead to user overload \texttt{few}.

While our dashboard layout is vitally important into making sure that the town leaders are using the dashboards, we should discuss the dashboard components that we used along with the philosophy of those key components. Starting with our parallel coordinate plot, which will allow for objective measures related to the QoL Survey. As our primary objective is to provide users with a town-centric approach: their town is at the center of our application, and comparisons to other, similar towns are secondary. We believe that this will allow the town leaders the ability to see as much data as possible with relation to their towns and others. The framework mentioned in \texttt{lee},``How do People Make Sense of Unfamiliar Visualizations?: A Grounded Model of Novice’s Information Visualization Sensemaking,'' allowed our work to determine that a novice user seeing a PCP will be able to determine two basic components: Visual Object (textual objects and non-textual objects) and Frame (frame of content and frame of visual encoding). The PCP in this study were found to be promising for novice users, when comparing to the Treemap and Chord Diagram.

@@

\subsection{Guided Discovery Learning (GDL)}
We leverage the framework of Guided Discovery Learning (GDL) to guide the town leaders to make discoveries using our interactive visualization. This framework leverages hints, feedback, and other helpful information to guide users in interactive exploration \texttt{dedonno}.

The Data Visualization component with GDL principles is a single page which populates the information in maps related to necessary services, including directions and distance to the fire department, schools, post offices, and hospitals. Value Boxes populate vital statistics sections that have information about the town's QoL Metrics and financial metrics, followed by a parallel coordinate plot that allows the town to see five towns that are similarly based on the most common variables in the towns, such as similar economy or population size. The graphical representations including a parallel coordinate plot and geographic maps [Figure 1]. The Rural Smart EDA Dashboard map visuals were built using OSRM route functions in R to amplify the accuracy of the distances from necessary services in town-centric point.

@@

\subsection{User Testing}
In this paper, we present the results from testing this application with users, and examine the implications for the feedback we received on our application design and on the more general design of dashboards for statistical novices. We examine which aspects of dashboard design ``common knowledge'' hold up in practice, and assess why these principles may not hold in our use-case.

Complex visualizations are difficult to assess in vivo: it is fairly easy to design an experiment to test a single graphic, but much harder to test the interactive dashboards which consist of systems of displays and user interactions. As a result, it is important to continuously assess which ``common knowledge'' design principles stemming from more controlled experiments hold up in real-world situations.

Using the framework from "Creating design guidelines for building city dashboards from a user's perspectives," \texttt{young}, we will help summarize a users concerns utilizing six main critiques, which are framed into questions. We will adapt these questions to better help the town leaders communicate clear feedback for our team to use. We will collect how the users are experiencing the app with these set questions:
\begin{itemize}
\item Is the information the dashboard presents clear? \textit{(Scope)}
\item Is the dashboard easy to navigate? \textit{(Usability)}
\item Is the dashboard used in a socially responsible manner? \textit{(Ethics)}
\item Do you believe that you can trust the Rural Shrink Smart dashboard? \textit{(Validity)}
\item Do you find value in the Rural Shrink Smart dashboard? \textit{(Utility)}
\end{itemize}

\subsubsection{Beta Testing}
Our Beta testing of a design dashboard before deploying the product to our town leaders, we allowed users from the following groups:
\begin{itemize}
    \item Graphics Group Members (Users know Statistical Graphics)
    \item Rural Shrink Smart Project Members (Users know the project with little background in Statistical Graphics)
    \item  Others (Member of doctoral community that has background in educational adaptivity)
\end{itemize}

@@

\subsection{Visualization Revisited}
In our statistical graphs, we used Parallel Coordinate Plots (PCPs) to allow the user to look at a variety of variables, while the Guided Discovery Learning was used to makes sure that the users were informed of the information that was presented, it did not allow for the town leaders any more ease in understanding. As a result, we decided to look for easier ways to display data, such as boxplots and barplots. This direction will be limited but may be useful foor the sake of ease on the town leaders.

@@

\subsection{Feedback}
We will continue to engage with twon leaders to ensure that we are developing a product that will help them make better data-driven decisions that are useful to increasing their quality of life, which may lend itself to help less shrinkage in these rural areas.

>>>>

\documentclass[letterpaper,inpress]{jdsart}

\setcounter{page}{1}
\pubmonth{July}
\pubyear{2020}
\volume{xx}
\issue{xx}
\doi{0000}

\usepackage[utf8]{inputenc}
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage{amsfonts,amsmath,amssymb,amsthm} \usepackage{booktabs} \usepackage{lipsum}

\begin{document}
\begin{frontmatter}

\title{Exploring Rural Shrink Smart Through Guided Discovery Dashboards}

\author[1]{
  \inits{D.}
  \fnms{Denise}
  \snm{Bradford}  \thanksref{1}  \ead{denise.bradford@huskers.unl.edu}}
\author[1]{
  \inits{S.}
  \fnms{Susan}
  \snm{VanderPlas}  \thanksref{1}}

\thankstext[type=corresp,id=1]{Corresponding author}
\address[1]{Department of Statistics, 
  \institution{University of Nebraska - Lincoln}, \cny{United States of America}}

\begin{abstract}
Many small and rural places are shrinking. Interactive dashboards are the most common use cases for data visualization and context for exploratory data tools. In our paper, we will explore the specific scope of how dashboards are used in small and rural area to empower novice analysts to make data-driven decisions. Our framework will suggest a number of research directions to better support small and rural places from shrinking using an interactive dashboard design, implementation and use for the every day analyst.
\end{abstract}

\begin{keywords}
\kwd{Interactive Dashboards}\kwd{Exploratory Data Analysis (EDA)}\kwd{Parallel Coordinate Plots (PCP)}\kwd{Guided Discovery Learning (GDL)}.
\end{keywords}

\end{frontmatter}

\hypertarget{abstract}{%
\section{Abstract}\label{abstract}}

@@

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

@@

Data collected from the Index of Relative Rurality (IRR) \url{https://www.ers.usda.gov/topics/rural-economy-population/rural-classifications/} were used in the SCC dashboard to help classify the towns. The Index of Relative Rurality (IRR) is a continuous, threshold-free, and unit-free measure of rurality. It is an alternative to the traditional discrete threshold-based classifications.The IRR ranges between 0 (low level of rurality, i.e., urban) and 1 (most rural). Four steps are involved in its design:

\begin{enumerate}
\item Identifying the dimensions of rurality: population size, density, remoteness, and built-up area.
\item Selecting measureable variables to adequately represent each dimension:
    \begin{itemize}
        \item Size: logarithm of population size
        \item Density: logarithm of population density.
        \item Remoteness: network distance.
        \item Built-up area: urban area (as defined by the US Census Bureau) as a percentage of total land area.
    \end{itemize}
\item Re-scaling the variables onto bounded scales that range from 0 to 1.
\item Selecting a link function: unweighted average of the four re-scaled variable.
\end{enumerate}

@@

One of the interesting features of this assembled dataset is that missing data can be missing for multiple reasons: not all state data is complete, but data about certain services may also be missing because towns do not offer that service.
Thus, in addition to the usual challenges of working with real-world data that is ``messy'' in a variety of ways, we also have to contend with missing data that is missing due to the size of the community or the lack of services. This makes both visualization and statistical analysis more complicated.

\hypertarget{methods}{%
\section{Methods}\label{methods}}

@@

Our audience being small rural towns, traditionally, do not have access:

\begin{itemize}
     \item Data scientists 
      \item Ability to easily leverage data collected locally to support decisions 
\end{itemize}

Our research team will provide communities with data about services in small town Iowa in order to assist with developing strategies to improve quality of life for their residents amid shrinking populations.

While data is important for data-driven decisions, we need to help minimize the amount of cognitive load for the town-leaders to digest the amount of data will be conveyed, as a result we will create a central place for the excess of data to be conveyed in a single comprehensive visual. We will utilize Parallel Coordinate Plot (PCPs). Create a map feature that will allow for town leaders to understand the distance of essential QoL components (e.g.~Fire Dept, Schools, etc.). Lastly, we will add an overview of town general basics related to the seven QoL factors.

@@

Thus, our initial set of dashboard design principles is as follows:

\begin{itemize}
\item The town leaders are the focus audience; thus, the town itself should be the central focus of the app.
\item Facilitate comparisons with other towns in order to allow the user to explore other potential solutions to offering services that enhance resident quality of life.
\item Present the user with peer comparisons in order to widen the scope of exploration beyond the initial set of obvious peers in the local region.
\item Allow for more detailed data and feature requests to improve the dashboard design over time.
\end{itemize}

@@

\subsubsection{Dashboard Components}

The large set of publicly available data (primarily from \url{data.iowa.gov}) we have assembled is useful, but we must be careful with how we present this data because it would be easy to overwhelm the user with small details that mask the bigger picture. We select a small subset of towns (out of the 999 towns in Iowa) and a small subset of variables of interest to start with, and then allow the user to increase the complexity of the display in accordance with their interest. This avoids some of the pitfalls of dashboard design that can easily lead to user overload \texttt{few}.

While our dashboard layout is vitally important into making sure that the town leaders are using the dashboards, we should discuss the dashboard components that we used along with the philosophy of those key components. Starting with our parallel coordinate plot, which will allow for objective measures related to the QoL Survey. As our primary objective is to provide users with a town-centric approach: their town is at the center of our application, and comparisons to other, similar towns are secondary. We believe that this will allow the town leaders the ability to see as much data as possible with relation to their towns and others. The framework mentioned in \texttt{lee},``How do People Make Sense of Unfamiliar Visualizations?: A Grounded Model of Novice's Information Visualization Sensemaking,'\,' allowed our work to determine that a novice user seeing a PCP will be able to determine two basic components: Visual Object (textual objects and non-textual objects) and Frame (frame of content and frame of visual encoding). The PCP in this study were found to be promising for novice users, when comparing to the Treemap and Chord Diagram.

@@

\subsection{Guided Discovery Learning (GDL)}

We leverage the framework of Guided Discovery Learning (GDL) to guide the town leaders to make discoveries using our interactive visualization. This framework leverages hints, feedback, and other helpful information to guide users in interactive exploration \texttt{dedonno}.

The Data Visualization component with GDL principles is a single page which populates the information in maps related to necessary services, including directions and distance to the fire department, schools, post offices, and hospitals. Value Boxes populate vital statistics sections that have information about the town's QoL Metrics and financial metrics, followed by a parallel coordinate plot that allows the town to see five towns that are similarly based on the most common variables in the towns, such as similar economy or population size. The graphical representations including a parallel coordinate plot and geographic maps {[}Figure 1{]}. The Rural Smart EDA Dashboard map visuals were built using OSRM route functions in R to amplify the accuracy of the distances from necessary services in town-centric point.

@@

\subsection{User Testing}

In this paper, we present the results from testing this application with users, and examine the implications for the feedback we received on our application design and on the more general design of dashboards for statistical novices. We examine which aspects of dashboard design ``common knowledge'\,' hold up in practice, and assess why these principles may not hold in our use-case.

Complex visualizations are difficult to assess in vivo: it is fairly easy to design an experiment to test a single graphic, but much harder to test the interactive dashboards which consist of systems of displays and user interactions. As a result, it is important to continuously assess which ``common knowledge'\,' design principles stemming from more controlled experiments hold up in real-world situations.

Using the framework from ``Creating design guidelines for building city dashboards from a user's perspectives,'' \texttt{young}, we will help summarize a users concerns utilizing six main critiques, which are framed into questions. We will adapt these questions to better help the town leaders communicate clear feedback for our team to use. We will collect how the users are experiencing the app with these set questions:

\begin{itemize}
\item Is the information the dashboard presents clear? \textit{(Scope)}
\item Is the dashboard easy to navigate? \textit{(Usability)}
\item Is the dashboard used in a socially responsible manner? \textit{(Ethics)}
\item Do you believe that you can trust the Rural Shrink Smart dashboard? \textit{(Validity)}
\item Do you find value in the Rural Shrink Smart dashboard? \textit{(Utility)}
\end{itemize}

\subsubsection{Beta Testing}

Our Beta testing of a design dashboard before deploying the product to our town leaders, we allowed users from the following groups:

\begin{itemize}
    \item Graphics Group Members (Users know Statistical Graphics)
    \item Rural Shrink Smart Project Members (Users know the project with little background in Statistical Graphics)
    \item  Others (Member of doctoral community that has background in educational adaptivity)
\end{itemize}

@@

\subsection{Visualization Revisited}

In our statistical graphs, we used Parallel Coordinate Plots (PCPs) to allow the user to look at a variety of variables, while the Guided Discovery Learning was used to makes sure that the users were informed of the information that was presented, it did not allow for the town leaders any more ease in understanding. As a result, we decided to look for easier ways to display data, such as boxplots and barplots. This direction will be limited but may be useful foor the sake of ease on the town leaders.

@@

\subsection{Feedback}

We will continue to engage with twon leaders to ensure that we are developing a product that will help them make better data-driven decisions that are useful to increasing their quality of life, which may lend itself to help less shrinkage in these rural areas.

\bibliography{jdsart-sample.bib}
\bibliographystyle{jds}

\end{document}
